---
title: "Packet 3 - Tidying Data and More Visulatizations"
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
author: "Todd CadwalladerOlsker"
date: "*Last updated:* `r Sys.Date()`"
output:
  rmdformats::downcute:
    downcute_theme: "chaos"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, eval = FALSE)
library(rmdformats)
library(openintro)
library(tidyverse)
```

## Tidying Data

**Lightly read (https://r4ds.had.co.nz/tidy-data.html).**

So far, we've been using the `openintro` data sets, which are great in may ways because they have already been *tidied*. That is, the data has been collected into a frame in such a way that it is easy for us to use.

Many data sets take some work to tidy up. There may be variables we want to ignore, missing values we need to deal with, and so on. 

For this exercise, read [this article](https://web.archive.org/web/20160223192726/https://www.nytimes.com/2015/12/06/opinion/sunday/let-math-save-our-democracy.html). As an activity, I'd like to  recreate the mean-median difference for the 2012 Pennsylvania House election.

Start with this link: [download the results of the House elections from 1976-2020](https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/IG0UN2). You should download the .zip file (You'll need to provide your name, email, and institution to download the data.) The .zip file should contain a .csv file and an .md file, place both of these in the working directory for your R project. (I'll also put these files on Canvas for you.)

Now, let's import the file into R. You can do this by clicking on the file in RStudio, then choosing `Import Dataset...`. Before clicking import, notice that RStudio will just execute the R commands for you. Click `Import`, and spend a moment looking over the data set. Then, click on the "codebook" (the .md file) and `Preview` it for information on the variables of the data set. 

Now, what we want to do is to transform this data set so that we can calculate the share of the votes in each district in the 2012 Pennsylvania House race. For simplicity, I'll just count the Democrat and Republican votes in each district, and ignore independent and third-party candidates.

Run the code below one line at a time, and at each step, look at the new data set that we've just created. How do the various functions used here work?

```{r tidying}
house_2012 <- 
  X1976_2020_house %>% 
  filter(year==2012) %>% 
  mutate(dist_id = paste(year, state_po, district), .after = state_po)

house_2012_dem <- 
  house_2012 %>%
  filter(party == "DEMOCRAT") %>% 
  mutate(dem_votes = candidatevotes)

# Old Version: 
# house_2012_dem_reduced <- house_2012_dem[, c(4,22)]
# ----
# Revised Version:
house_2012_dem_reduced <- 
  house_2012_dem %>% 
  select(dist_id,dem_votes)

house_2012_dem_reduced <- 
  house_2012_dem_reduced %>%
  group_by(dist_id) %>%
  summarise(dem_votes = sum(dem_votes, na.rm = T))


house_2012_rep <- 
  house_2012 %>%
  filter(party == "REPUBLICAN") %>% 
  mutate(rep_votes = candidatevotes)

# Old Version:
# house_2012_rep_reduced <- house_2012_rep[, c(4,22)]
# ----
# Revised Version:
house_2012_rep_reduced <- 
  house_2012_rep %>% 
  select(dist_id,rep_votes)

house_2012_rep_reduced <- 
  house_2012_rep_reduced %>%
  group_by(dist_id) %>%
  summarise(rep_votes = sum(rep_votes, na.rm = T))

house_2012_reduced <- 
  full_join(house_2012_dem_reduced,house_2012_rep_reduced,by = "dist_id") %>% 
  arrange(dist_id)

house_2012_reduced <- 
  house_2012_reduced %>% 
  replace(is.na(.),0)

house_2012_reduced <- 
  house_2012_reduced %>% 
  mutate(total_votes = dem_votes + rep_votes) %>% # Note: Easier code here
  mutate(pct_dem_votes = dem_votes / total_votes * 100) %>% 
  mutate(pct_rep_votes = rep_votes / total_votes * 100)

house_2012_PA <- house_2012_reduced %>% 
  filter(grepl("PA", dist_id))

median_dem_pct_PA <- median(house_2012_PA$pct_dem_votes)
mean_dem_pct_PA <- mean(house_2012_PA$pct_dem_votes)
```

## More Visualizations

Now, let's make some more plots. We can make a histogram again:

```{r histogram}
statsholder <- data.frame(stat = c("Median",
                                     "Mean"),
                          value = c(median_dem_pct_PA,
                                     mean_dem_pct_PA))

house_2012_PA %>% 
  ggplot(aes(x = pct_dem_votes)) +
  scale_color_brewer(palette = "Dark2") +
  geom_histogram(binwidth=1, color = "black", fill = "blue") + 
  geom_vline(data=statsholder ,aes(xintercept = value,
                               linetype = stat,
                               color = stat),size=1) +
  labs(title = "Percentage of Democratic votes by district",
       subtitle = "2012 Pennsylvania House race", 
       y = NULL,
       x = "Percentage of Democratic votes")
```

Try changing the binwidth up a bit to make this picture prettier. Alternatively, try a `dotplot`:

```{r dotplot}
house_2012_PA %>% 
  ggplot(aes(x = pct_dem_votes)) +
  geom_dotplot(color = "black", fill = "blue") + 
  geom_vline(data=statsholder ,aes(xintercept = value,
                               linetype = stat,
                               color = stat),size=1) +
  labs(title = "Percentage of Democratic votes by district",
       subtitle = "2012 Pennsylvania House race", 
       y = NULL,
       x = "Percentage of Democratic votes") +
  scale_y_continuous(breaks = NULL, labels = NULL) +
  scale_x_continuous(breaks = seq(20,95,by = 5)) +
  scale_color_brewer(palette = "Dark2")
```

We can spread the dots out a bit by messing with the binwidth value:

```{r dotplot2}
house_2012_PA %>% 
  ggplot(aes(x = pct_dem_votes)) +
  geom_dotplot(binwidth = 0.05, dotsize= 30, color = "black", fill = "blue") + 
  geom_vline(data=statsholder ,aes(xintercept = value,
                               linetype = stat,
                               color = stat),size=1) +
  labs(title = "Percentage of Democratic votes by district",
       subtitle = "2012 Pennsylvania House race", 
       y = NULL,
       x = "Percentage of Democratic votes") +
  scale_y_continuous(breaks = NULL, labels = NULL) +
  scale_x_continuous(breaks = seq(20,95,by = 5)) +
  scale_color_brewer(palette = "Dark2")
```

## To Do: 

For this assignment, we want to explore the question, "Has my local area gotten hotter over the past 20 years?" In other words, we are looking at "local warming" (rather than global warming). We'll explore this question in a couple of different ways.

Go to (https://www.ncdc.noaa.gov/cdo-web/datasets) and download a "Global Summary of the Month" data set. You can look at a station near where you live, or maybe another location that has some significance to you. Select data that goes back at least 20 years, longer if possible, and search for a station with good data. (Airports are usually good for weather data.) When you check out your data, you'll be allowed to select variables: Choose variables for maximum temperature (TMAX), number of days with max temp. above 70 (DX70), and number of days with max temp above 90 (DX90). You can also select other variables if you are interested. Be sure to download your data as a .csv file.

You'll need to provide your email address, and the website will send you a link to download your data. 

Once you have your data in hand, use R to tidy and transform your data. Your .pdf file should contain all of the following. For this assignment, *do not explain what you are seeing/doing in each step individually.* Instead, I want you to tell the story of the data at the end of the assignment.

a. Add a column (variable) that contains the month the data was collected, and another column that contains the year the data was collected (in other words , separate the DATE into "MONTH" and "YEAR".) See the `zoo` and `lubridate` packages for help on doing this.

b. Create stacked boxplots for the TMAX variable, separated by month. Normally, it would be messy to have this many stacked boxplots, but don't worry about it here.

c. Create stacked boxplots for the TMAX variable, separated by year (Hint: you may need to use `factor(YEAR)` in your ggplot instead of just YEAR). 

d. Create a scatterplot with YEAR on the x-axis, and TMAX on the y-axis.

e. Create a *line plot* (use `geom_line` instead of `geom_point`) with YEAR on the x-axis, TMAX on the y-axis, but using data only from February of each year. (You can also show both points *and* lines by including both `geom_point` and `geom_line` in the same `ggplot`.) 

f. Make a new data frame that looks year-by-year only. That is, your data frame should contain one row for each year, and contains the maximum temperate for the year, the total number of days above 70 for the year, and the total number of days above 90 for the year.

g. Using this new data frame, make a line plot with YEAR on the x-axis, and either DX70 or DX90 on the y-axis (the choice will depend on how hot the weather typically is at your chosen weather station.)

h. Finally, tell the story of your data. Does the maximum temperature for the year appear to be increasing over time? Does February seem to be hotter now than it was 20 years ago? Does there seem to have more hot days (over 70, or over 90) than there were 20 years ago? Remember that for now, we are just exploring the data, so just looking at the plots (without any further analysis) is fine.

## Reading Assignment:

Read [this editorial from the Journal of Statistics and Data Science Education.](https://www.tandfonline.com/doi/full/10.1080/26939169.2021.1906555). Write a short response around the question, "What classroom activities could I do differently in light of diversity, equity, and inclusion issues?" Give at least one specific example of an activity you have used, if possible.
